---
title: "Instacart EDA"
author: "Mark Rossi"
date: "June 28, 2017"
output: 
  html_document:
    theme: cosmo
---

### Hello! This notebook is an initial exploratory analysis to support the Instacart Basket Challenge. 

Instacart combines a grocery-ordering app with a delivery service: users select products on the application, then personal shoppers do the shopping and deliver it to their doorstep. 

The underlying data science supporting this service uses transactional data to provide recommendations to a user along the way during their session. 

For this challenge specifically, the task is to predict which previously purchased products will be in a user's next order. 

### Import libraries & get our data
```{r message=FALSE, warning=FALSE, results='hide'}
library(data.table)
library(dplyr)
library(ggplot2)
library(knitr)
library(stringr)
library(DT)

orders <- fread('Data/orders.csv')
products <- fread('Data/products.csv')
order_products <- fread('Data/order_products__train.csv')
order_products_prior <- fread('Data/order_products__prior.csv')
aisles <- fread('Data/aisles.csv')
departments <- fread('Data/departments.csv')
```

```{r include=FALSE}
options(tibble.width = Inf)
```

### Data {.tabset}

#### orders

```{r}
kable(head(orders, 12))
glimpse(orders)
```

#### order_products_train
```{r, result='asis'}
kable(head(order_products, 12))
glimpse(order_products)
```

#### products
```{r, result='asis'}
kable(head(products, 12))
glimpse(products)
```

#### order_products_prior
```{r, result='asis'}
kable(head(order_products_prior, 12))
glimpse(order_products_prior)
```

#### aisles
```{r, result='asis'}
kable(head(aisles, 12))
glimpse(aisles)
```

#### departments
```{r, result='asis'}
kable(head(departments, 12))
glimpse(departments)
```

### Some basic recoding:
```{r message=FALSE, warning=FALSE}
# orders: make sure hour of day and is numeric and eval_set is made a factor
orders <- orders %>% mutate(order_hour_of_day = as.numeric(order_hour_of_day), eval_set = as.factor(eval_set))

# products: set product_name to factor
products <- products %>% mutate(product_name = as.factor(product_name))

# same for names of aisles & departments
aisles <- aisles %>% mutate(aisle = as.factor(aisle))
departments <- departments %>% mutate(department = as.factor(department))
```


### Let's look at the orders themselves first...

#### Hour of day
Unsurprisingly, orders seem pretty concentrated between 10am and 4pm
```{r warning=FALSE}
orders %>% 
  ggplot(aes(x=order_hour_of_day)) + 
  geom_histogram(stat="count",fill="blue")
```

#### Day of week

```{r warning=FALSE}
orders %>% 
  ggplot(aes(x=order_dow)) + 
  geom_histogram(stat="count",fill="blue")
```

#### How do DOW/hour of day compare?
This one is sort of interesting. Orders tend to 
```{r warning=FALSE}
orders %>%
  group_by(order_dow, order_hour_of_day) %>%
  summarize(num_orders = n()) %>%
  ggplot(aes(order_dow, order_hour_of_day)) +
  geom_tile(aes(fill=num_orders)) +
  scale_fill_continuous(low="white", high="steelblue")
```

#### Typical # days between orders?
```{r warning=FALSE}
orders %>% 
  ggplot(aes(x=days_since_prior_order)) + 
  geom_histogram(stat="count", fill="blue")
```

#### How many prior orders are there?
```{r warning=FALSE}
orders %>% filter(eval_set=="prior") %>% count(order_number) %>% ggplot(aes(x=order_number, y=n)) + geom_line(color="red", size=1) + geom_point(size=2, color="blue")
```

### How many items do people buy? {.tabset}

#### Training set
```{r warning=FALSE}
order_products %>%
  group_by(order_id) %>%
  summarize(n_items = last(add_to_cart_order)) %>%
  ggplot(aes(x=n_items)) +
  geom_histogram(stat="count", fill="red") + 
  geom_rug() + 
  coord_cartesian(xlim=c(0,80))
```

#### Prior orders set
```{r warning=FALSE}
order_products_prior %>%
  group_by(order_id) %>%
  summarize(n_items = last(add_to_cart_order)) %>%
  ggplot(aes(x=n_items)) +
  geom_histogram(stat="count", fill="red") + 
  geom_rug() + 
  coord_cartesian(xlim=c(0,80))
```

### Looking more closely now at the items themselves...

#### What are the 10 most popular items on the site?
```{r fig.height=5.5}
tmp <- order_products %>% 
  group_by(product_id) %>%
             summarize(count = n()) %>%
             top_n(10, wt=count) %>%
             left_join(select(products, product_id, product_name), by="product_id") %>%
                         arrange(desc(count))

kable(tmp)

tmp %>% 
  ggplot(aes(x=reorder(product_name,-count), y=count))+
  geom_bar(stat="identity",fill="red")+
  theme(axis.text.x=element_text(angle=90, hjust=1),axis.title.x = element_blank())

```

### What about reorders?

#### How often is an item a reorder versus a 'fresh' one?
```{r warning=FALSE, fig.width=4}
tmp <- order_products %>%
  group_by(reordered) %>%
  summarize(count = n()) %>%
  mutate(reordered = as.factor(reordered)) %>%
  mutate(proportion = count/sum(count))

kable(tmp)

tmp %>%
  ggplot(aes(x=reordered, y=count, fill=reordered)) +
  geom_bar(stat="identity")
```

#### For individual orders, what is a typical ratio of reordered items to new ones?
```{r warning = FALSE}
order_products %>% group_by(order_id) %>% summarize(reorder_proportion = round((sum(reordered) / n()) / 0.05) * 0.05) %>% mutate(reorder_proportion = as.factor(reorder_proportion)) %>% ggplot(aes(x=reorder_proportion)) + geom_histogram(stat="count")
```

#### At what point in the shopping process do folks typically add reordered items to their carts?
```{r warning = FALSE}
order_products %>% group_by(add_to_cart_order) %>% summarize(reorder_proportion = sum(reordered) / n()) %>% ggplot(aes(x=add_to_cart_order, y=reorder_proportion)) + geom_bar(stat="identity")
```



#### Which items are REordered the most? 

```{r warning=FALSE, fig.height=5.5}
tmp <- order_products %>%
  group_by(product_id) %>%
  summarize(proportion_reordered = mean(reordered), n=n()) %>%
  filter(n>40) %>%
  top_n(10, wt=proportion_reordered) %>%
  arrange(desc(proportion_reordered)) %>%
  left_join(products, by="product_id")

kable(tmp)

tmp %>% 
  ggplot(aes(x=reorder(product_name,-proportion_reordered), y=proportion_reordered))+
  geom_bar(stat="identity",fill="red")+
  theme(axis.text.x=element_text(angle=90, hjust=1),axis.title.x = element_blank())+coord_cartesian(ylim=c(0.85,0.95))
```



